Indel Historian: fast and accurate ancestral sequence reconstruction

### 2016-10-18 doc/main.tex
Variable branch lengths

Multiple alignments are used for several purposes in bioinformatics,
only one of which is homology-directed protein structure prediction,
yet structurally-derived alignments have tended to dominate benchmarks.
Alignment accuracy scores (such as SPS, TCS, AMA, and the Cline shift score) quantify the number of correctly aligned residues
compared to a benchmark dataset of structurally-informed reference alignments \citep{ThompsonEtAl2005}.
Techniques that score highly in these tests include using Bayesian decision theory to maximize the expected
alignment accuracy \citep{NotredameEtAl2000,DoEtAl2005,SchwartzPachter2007,BradleyEtAl2009}
and fine-tuning the alignment scoring function for contextual details, such as the reduced propensity for indels in
hydrophobic protein regions \citep{Edgar2004b,KatohEtAl2005,LarkinEtAl2007}.
These have been accompanied by performance optimizations of 
rate-limiting steps such as all-versus-all pairwise sequence comparison \citep{Edgar2004b,BradleyEtAl2009}.

There is evidence that for evolutionary applications, such as reconstructing trees or ancestral sequences,
different alignment tools (or at least different tool-assessment metrics) might be preferable.
Empirical studies suggest that, for the purposes of estimating
molecular evolutionary parameters---such as indel rates \citep{Westesson2012-zg},
dN/dS ratios \citep{Redelings2014},
or trees \citep{LoytynojaGoldman2008}---it is advantageous to use an explicit statistical model of the sequence evolution process
and to treat alignment rigorously as a ``missing data'' problem.
One possible explanation is that, for protein structure prediction,
detailed reconstruction of indel histories in fast-evolving regions (such as loops) is unnecessary
(as structurally homologous loops can simply be aligned),
whereas evolutionary analyses can make use of this information.

One previous study sought to quantify systematic biases introduced into the
estimation of indel rates \citep{Westesson2012-zg},
using the evolution simulator indel-Seq-Gen \citep{StropeEtAl2009}.
The ProtPal program \citep{Westesson2012-zg}, which models indel evolution using 
transducers---finite-state machines which can be multiplied together like substitution matrices \citep{BouchardCote2013}---introduced
the least biases of the tools evaluated (RMS error 25\% for insertion rates and 25\% for deletion rates),
followed by PRANK (RMS error 40\% for insertions, 34\% for deletions) and then a MUSCLE-based workflow (45\%, 37\%).
Biases were substantially worse at higher indel rates: e.g. for the highest simulated insertion rate of 0.08 insertions/substitution,
the RMS estimation errors were 43\% (ProtPal), 64\% (PRANK) and 74\% (MUSCLE).

>>>>

Multiple alignments are used for several purposes in bioinformatics,
only one of which is homology-directed protein structure prediction,
yet structurally-derived alignments have tended to dominate benchmarks.
There is evidence that for evolutionary applications, such as reconstructing trees or ancestral sequences,
different alignment tools (along with different tool-assessment metrics) might be preferable.
Aligners that are optimized for detecting structural homology may not do so well at recovering information about substitution rates,
whereas explicit statistical models of the sequence evolution process may do a better job at estimating these parameters.

Empirical studies suggest that, for the purposes of estimating
molecular evolutionary parameters---such as indel rates \citep{Westesson2012-zg},
dN/dS ratios \citep{Redelings2014},
or trees \citep{LoytynojaGoldman2008}---it is advantageous to use a statistical model of evolution
and to treat alignment rigorously as a ``missing data'' problem.
One possible explanation is that, for protein structure prediction,
detailed reconstruction of indel histories in fast-evolving regions (such as loops) is unnecessary
(as structurally homologous loops can simply be aligned),
whereas evolutionary analyses can make use of this information.

The benchmarks commonly used to evaluate aligners use accuracy scores
(such as SPS, TCS, AMA, and the Cline shift score)
which quantify the number of correctly aligned residues
compared to a dataset of structurally-informed reference alignments \citep{ThompsonEtAl2005}.
Techniques that score highly in these tests include using Bayesian decision theory to maximize the expected
alignment accuracy \citep{NotredameEtAl2000,DoEtAl2005,SchwartzPachter2007,BradleyEtAl2009}
and fine-tuning the alignment scoring function for contextual details, such as the reduced propensity for indels in
hydrophobic protein regions \citep{Edgar2004b,KatohEtAl2005,LarkinEtAl2007}.
Aligners have also been tuned for performance by optimizing rate-limiting steps
uch as all-versus-all pairwise sequence comparison \citep{Edgar2004b,BradleyEtAl2009}.

In a previous study, we sought to quantify systematic biases introduced into the
estimation of indel rates \citep{Westesson2012-zg},
using the evolution simulator indel-Seq-Gen \citep{StropeEtAl2009}.
The ProtPal program \citep{Westesson2012-zg}, which models indel evolution using 
transducers---finite-state machines which can be multiplied together like substitution matrices \citep{BouchardCote2013}---introduced
the least biases of the tools evaluated (RMS error 25\% for insertion rates and 25\% for deletion rates),
followed by PRANK (RMS error 40\% for insertions, 34\% for deletions) and then a MUSCLE-based workflow (45\%, 37\%).
Biases were substantially worse at higher indel rates: e.g. for the highest simulated insertion rate of 0.08 insertions/substitution,
the RMS estimation errors were 43\% (ProtPal), 64\% (PRANK) and 74\% (MUSCLE).
